---
title: "CNVfilteR: Remove false positives of CNV calling tools by using SNV calls"
author: "Jose Marcos Moreno-Cabrera <jmoreno@igtp.cat> and Bernat Gel <bgel@igtp.cat> "
date: "`r doc_date()`"
package: "`r pkg_ver('CNVfilteR')`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{CNVfilteR vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, include=FALSE}
library(knitr)
opts_chunk$set(concordance=FALSE)
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 5)
set.seed(21666641)
```


# Introduction

Many tools for germline copy number variant (CNV) detection from 
NGS data have been developed. Usually, these tools were designed
for different input data like WGS, WES or
panel data, and their performance may depend on the CNV size. Available
benchmarks show that all these tools obtain false positives, sometimes 
reaching a very high number of them.

With the aim of reducing the number of false positives,
`r BiocStyle::Biocpkg("CNVfilteR")` identifies those germline CNVs that can 
be discarded. This task is performed by using the germline 
single nucleotide variant (SNV) calls that are usually obtained in 
common NGS pipelines. As VCF field interpretation is key when working
with these files, `r BiocStyle::Biocpkg("CNVfilteR")` specifically supports 
VCFs produced by VarScan2, Strelka/Strelka2, freeBayes, HaplotypeCaller (GATK),
and UnifiedGenotyper (GATK). Additionally, results can be plotted using the 
functions provided by the R/Bioconductor packages
[karyoploteR](http://bioconductor.org/packages/karyoploteR/) and 
[CopyNumberPlots](http://bioconductor.org/packages/CopyNumberPlots/).



# Installation

`r BiocStyle::Biocpkg("CNVfilteR")` is a 
[Bioconductor](http://bioconductor.org) package and to install it we have
to use `r BiocStyle::Biocpkg("BiocManager")`.

```{r getPackage, eval=FALSE}
  if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager")
  BiocManager::install("CNVfilteR")
```


We can also install the package from github to get the latest **devel version**.

```` {r, eval = FALSE}
  BiocManager::install("jpuntomarcos/CNVfilteR")
```


# Quick Start

Below we show a full example that covers the usual steps: CNVs data loading,
SNVs loading, identifying false postives and plotting the results.

First, we can load some CNV tool results:
```{r message=FALSE}
library(CNVfilteR)

cnvs.file <- system.file("extdata", "DECoN.CNVcalls.csv", package = "CNVfilteR", mustWork = TRUE)
cnvs.gr <- loadCNVcalls(cnvs.file = cnvs.file, chr.column = "Chromosome", start.column = "Start", end.column = "End", cnv.column = "CNV.type", sample.column = "Sample", genome = "hg19")

```

Then, we load the SNVs stored in a couple of VCF files. 
```{r }
vcf.files <- c(system.file("extdata", "variants.sample1.vcf.gz", package = "CNVfilteR", mustWork = TRUE),
               system.file("extdata", "variants.sample2.vcf.gz", package = "CNVfilteR", mustWork = TRUE))
vcfs <- loadVCFs(vcf.files, cnvs.gr = cnvs.gr)
```
We observe that the function recognized VarScan2 as the source, so fields
were selected and allele frequency consequently. Now we can call 
`filterCNVs()` to identify those CNVs that can be discarded.

```{r message=FALSE }
results <- filterCNVs(cnvs.gr, vcfs)
names(results)
```

And we can check those CNVs that can be filtered out:

```{r}
results$cnvs[results$cnvs$filter == TRUE]
```

As an example, we can observe that the CNV with `cnv.id`=3 contains 4 variants:
2 in favor of discarding it, two against discarding it. If we want 
to know more about the variants falling in a certain CNV we can do:
```{r}
results$variantsForEachCNV[["3"]]
```

Two variants are close to the default expected heterozygous frequency,
0.5, so they obtain a positive score. The other two variants are not 
so clearly close to the default expected duplication value, 0.33, so
they obtain a low negative score. All these default values and others can
be modified in the `filterCNVs()` function.

Finally, we may be interested in plotting the results. For example, we can
plot easily the duplication CNV with `cnv.id`=3 and all the variants falling in
it.

```{r, fig.wide=TRUE}
plotVariantsForCNV(results, "3")
```

We can do the same to plot the deletion CNV with `cnv.id`=19, where all 
variants discard the CNV except one homozygous variant that does not give us any 
information for supporting or discarding the CNV:

```{r, fig.wide=TRUE}
plotVariantsForCNV(results, "19")
```


On the opposite side, we can observe those CNVs that cannot be discarded:
```{r}
results$cnvs[results$cnvs$filter != TRUE]
```

For example, the CNV with `cnv.id`=14 contains one variant. If we get
the variant info, we see that the variant has an allele frequency very close to
the default expected duplication value, 0.66.

```{r}
results$variantsForEachCNV[["14"]]
```

So, no evidence was found for discarding the CNV. We can also plot the CNV and 
the variant:
```{r, fig.wide=TRUE}
plotVariantsForCNV(results, "3")
```




# Loading Copy-Number Data

`r BiocStyle::Biocpkg("CNVfilteR")` functions expect germline CNVs calls to be a
`GRanges` object with a few specificic metadata columns: 

* **cnv** for storing the CNV type: duplication or deletion.
* **sample** for storing the sample name of the CNV.
  
You can create this object yourself, but 
`r BiocStyle::Biocpkg("CNVfilter")` provides the proper function to do this, 
`loadCNVcalls()`. This function can interpret any tsv o csv file by indicating
which columns store the information. For example, in the following code, the
`chr.column` column is stored in the "Chromosome" column of the `cnvs.file`.
```{r message=FALSE}

cnvs.file <- system.file("extdata", "DECoN.CNVcalls.csv", package = "CNVfilteR", mustWork = TRUE)
cnvs.gr <- loadCNVcalls(cnvs.file = cnvs.file, chr.column = "Chromosome", start.column = "Start", end.column = "End", cnv.column = "CNV.type", sample.column = "Sample", genome = "hg19")

```

`loadCNVcalls()` can interpret different types of CNVs. Among other options, 
separator can be selected using the `sep` parameter (defaults to *\\t*),
and first lines can be skipped using the `skip` parameter (defaults to 0). Also,
the value used in `cnv.column` to store the CNV type can be modified
using the `deletion` and `duplication` parameters (defaults to "deletion" and 
"duplication", respectively). If, for example, our `cnv.column` uses "CN1" and
"CN3" for deletion and duplication (respectively), we should indicate 
the function to use these values:
```{r eval=FALSE}
cnvs.gr.2 <- loadCNVcalls(cnvs.file = cnvs.file.2, deletion = "CN1", duplication = "CN3", chr.column = "Chromosome", start.column = "Start", end.column = "End", cnv.column = "CNV.type", sample.column = "Sample")
```


Some CNV tools generate results where the CNV location is stored in a single
column with the format *chr:start-end* (i.e. *1:538001-540000*). In this
case, we can call `loadCNVcalls()` using the `coord.column` instead of the 
`chr.column`, `start.column` and `end.column` columns.



# Loading Variants Data

Common NGS pipelines produce germline variant calling (SNVs or INDELs)
in a VCF file. However, VCF interpretation is 
challenging due to the flexibility provided by the VCF format definition. 
It is not straightforward to interpret correctly the fields in the VCF file 
and compute the allele frequency. `loadVCFs()` interprets automatically 
VCFs produced by VarScan2, Strelka/Strelka2, freeBayes, HaplotypeCaller (GATK),
and UnifiedGenotyper (GATK).

In the following example the function recognizes VarScan2 as the source.
```{r }
vcf.files <- c(system.file("extdata", "variants.sample1.vcf.gz", package = "CNVfilteR", mustWork = TRUE),
               system.file("extdata", "variants.sample2.vcf.gz", package = "CNVfilteR", mustWork = TRUE))
vcfs <- loadVCFs(vcf.files, cnvs.gr = cnvs.gr)
```

We can also load the VCF file spicifying how to interpret it, which can be
useful if the VCF was generated by a caller not supported by 
`r BiocStyle::Biocpkg("CNVfilteR")`. For example we can specify the ref/alt 
fields:
```{r }
vcfs <- loadVCFs(vcf.files, cnvs.gr = cnvs.gr, vcf.source = "MyCaller", ref.support.field = "RD", alt.support.field = "AD")
```

Alternatively, we can set the `list.support.field` parameter so that field
will be loaded assuming that it is a list in this order: reference allele, 
alternative allele. As an example:
```{r}
vcf.file3 <- c(system.file("extdata", "variants.sample3.vcf", package = "CNVfilteR", mustWork = TRUE))
vcfs3 <- loadVCFs(vcf.file3, cnvs.gr = cnvs.gr, vcf.source = "MyCaller", list.support.field = "AD")
```

## VCF free of artifacts

As VCF will be used for determining if a CNV can be filtered out, **we
recommend to use a VCF file free of false-positive SNVs** (as possible)
to improve the accuracy when identifying those CNVs. The aim is to discard the
SNVs that are probably false-positives and also those which allele 
frequency is no reliable.

### Minimun total depth

`loadVCFs` will load only variants with a **minimum total depth** equal or 
greater than `min.total.depth`. The default value, 10, showed good performance
on our CNVfilteR evaluation on 3 WGS samples (data will be available soon) 
and using Strelka2 as SNV variant caller. This value **should be
adapted to your variant calling tool and sequencing depth conditions**. For 
example, we used a `min.total.depth`
of 30 when using CNVfilteR on panel (targeted-enrinched) samples with high 
coverage and VarScan2 as SNV caller.


### Regions to exclude

Additionally, we can exclude
those complex regions that have already known alignement artifacts 
with the parameter `regions.to.exclude`. 
In this example, we exclude PMS2, PRSS1, and
FANCD2 genes because they are pseudogenes with alignments artifacts:
```{r}
regions.to.exclude <- GRanges(seqnames = c("chr3","chr7", "chr7"), ranges = IRanges(c(10068098, 6012870, 142457319), c(10143614, 6048756, 142460923)))
vcfs4 <- loadVCFs(vcf.files, cnvs.gr = cnvs.gr, regions.to.exclude = regions.to.exclude)
```

### INDELs excluded by default

Also, the parameter `exclude.indels` indicates whether to exclude INDELs when 
loading the variants. TRUE is the default and **recommended** value given 
that INDELs allele frequency varies differently than SNVs. Including
INDELs may allow the algorithm to identify more CNVs to discard with a greater
risk of identifying them wrongly. Additionally, any SNV overlapping an INDEL
will be ignored because the SNV allele frequency may be affected in that region.


## Other settings

The function `loadVCFs()` also adapts to different needs. If `sample.names` parameter is
not provided, the sample names included in the VCF itself will be used. 
Both single-sample and multi-sample VCFs are accepted, but when 
multi-sample VCFs are used, `sample.names` parameter must be NULL.

If VCF is not compressed with bgzip, the function compresses it and generates
the .gz file. If .tbi file does not exist for a given VCF file, the function
also generates it. All files are generated in a temporary folder.

See `loadVCFs()` documentation to see other parameters info.



# Identifying false positives

The task of identifying false positives is performed by the `filterCNVs()`
function. It checks all the variants (SNVs and optionally INDELs) falling in
each CNV present in `cnvs.gr` to identify those CNVs that can be filtered out. It
returns an S3 object with 3 elements: `cnvs`, `variantsForEachCNV` and 
`filterParameters`:

```{r}
results <- filterCNVs(cnvs.gr, vcfs)
tail(results$cnvs)
```


Observe that **those CNVs that can be filtered out have the value TRUE in the 
column `filter`**. `r BiocStyle::Biocpkg("CNVfilteR")` employs two
different strategies for identifying those CNVs:

 * A **deletion CNV** can be filtered out if there is at least 
 `ht.deletions.threshold`% of heterozygous variants in
 the CNV. Default `ht.deletions.threshold` value is 30, so 30% is required.
 * A **duplication CNV** can be filtered out if the total `score` is >= 
 `dup.threshold.score` after computing all 
 heterozygous variants falling in that CNV. Default `dup.threshold.score` value
 is 0.5. How the score is computed for each variant is explained in 
 the next section.


## Scoring model for duplication CNVs 

The scoring model for determining whether a certain duplication CNV can be 
discarded
is based on the allele frequency for each heterozygous variant falling in
that CNV:

 * In common conditions with no presence of a duplication CNV, the allele 
 frequency of a heterozygous variant is expected to be close to 50% 
 (`expected.ht.mean`). So, a variant with an allele frequency close to 50%
 gives us evidence of the non-existence of a duplication CNV, so the CNV could 
 be discarded.
 * On the opposite side, if the variant occurs in the same region of a 
certain duplication CNV, the allele frequency is expected to be close to 33.3%
(`expected.dup.ht.mean1`) when the variant **is not** in the same allele 
 than the duplication CNV, and 66.6% (`expected.dup.ht.mean2`) when the variant 
 **is** in the same allele than 
 the duplication CNV call. So, a variant with an allele frequency close to 
 33.3% or 66.6% gives us evidence of the existence of duplication CNV.



Although we can expect that most of the variants are close to the expected means
(33.3%, 50%, and 66.6%), many of them can be far from any expected mean. The 
scoring model implemented in the `filterCNVs()` function measures
the evidence - for discarding a certain CNV - using
a scoring model. The scoring model is based on the fuzzy logic, where elements 
can have any value between 1 (True) and 0 (False). Following this idea, 
**each variant will be scored with a value between 0 and 1 depending on 
how close is the allele frequency to the nearest expected mean**. 

 * For those variants with an allele frequency close to the expected mean when
 no duplication CNV occurs (defaults 50%),
 the score will be positive in the interval [0, 1].
 * For those variants with an allele frequency close to the expected mean when
 a duplication CNV occurs (defaults 33.3%, 66.6%), the score will be negative 
 in the interval [-1, 0].

The total `score` is computed among all the variants falling in the CNV. If the 
`score` is greater than the `dup.threshold.score`, the CNV can be discarded.

A common way of applying the fuzzy logic is using the sigmoid function. 
CNVfilteR uses the sigmoid function implemented in the
[pracma](https://cran.r-project.org/web/packages/pracma/index.html) package, 
which is defined as
$$
\begin{aligned}
 y = 1/(1 + e^{-c1(x−c2)})
\end{aligned}
$$

The scoring model is built on 6 sigmoids defined on 6 different intervals. The
c1 parameter is 2 by default (`sigmoid.c1`), and the c2 parameter 
is defined for the 6 sigmoids (`sigmoid.c2.vector`).

 * First sigmoid: interval [20, `expected.dup.ht.mean1`], c2=28
 * Second sigmoid: interval [`expected.dup.ht.mean1`, `sigmoid.int1`], c2=38.3
 * Third sigmoid: interval [`sigmoid.int1`, `expected.ht.mean`], c2=44.7
 * Fourth sigmoid: interval [`expected.ht.mean`, `sigmoid.int2`], c2=55.3
 * Fifth sigmoid: interval [`sigmoid.int2`, `expected.dup.ht.mean2`], c`=61.3
 * Sixth sigmoid: interval [`expected.dup.ht.mean2`, 80], c2=71.3
 
Where `sigmoid.int1` is the mean between `expected.dup.ht.mean1` and 
`expected.ht.mean`, and `sigmoid.int2` is the mean between 
`expected.dup.ht.mean2` and `expected.ht.mean`.
 
The scoring model can be plotted using the `plotScoringModel()` function.
```{r, fig.height=6, fig.wide=TRUE}
p <- results$filterParameters
plotScoringModel(expected.ht.mean = p$expected.ht.mean, 
                 expected.dup.ht.mean1 = p$expected.dup.ht.mean1,
                 expected.dup.ht.mean2 = p$expected.dup.ht.mean2,
                 sigmoid.c1 = p$sigmoid.c1, 
                 sigmoid.c2.vector = p$sigmoid.c2.vector)

```

And the scoring model can be modified when calling the `filterCNVs()` function. 
Let's see how the model changes when we modify the parameter `sigmoid.c1` 
(1 instead of 2): 
```{r, fig.height=6, fig.wide=TRUE}
plotScoringModel(expected.ht.mean = p$expected.ht.mean, 
                 expected.dup.ht.mean1 = p$expected.dup.ht.mean1,
                 expected.dup.ht.mean2 = p$expected.dup.ht.mean2,
                 sigmoid.c1 = 1, 
                 sigmoid.c2.vector = p$sigmoid.c2.vector)
```

We can also modify the `sigmoid.c2.vector` parameter for each sigmoid function. For example, to 
make the central sigmoids narrower:
```{r, fig.height=6, fig.wide=TRUE}
plotScoringModel(expected.ht.mean = p$expected.ht.mean, 
                 expected.dup.ht.mean1 = p$expected.dup.ht.mean1,
                 expected.dup.ht.mean2 = p$expected.dup.ht.mean2,
                 sigmoid.c1 = p$sigmoid.c1, 
                 sigmoid.c2.vector = c(28, 38.3, 46.7, 53.3, 61.3, 71.3))
```

## The margin.pct parameter
The idea under this parameter is to ignore the variants placed in the "margins"
of the CNV given that WGS CNV callers make unaccurate calls: they work on 
windows sizes that may not correspond with the real CNV size. In consequence, we 
prefer to exlude SNVs falling in the margins to avoid potential "noise" (SNVs
out of the real CNV). In our evaluation on the HuRef WGS sample, we observed that
values between 10 (default value) and 20 performed the best (data available soon).

Summarizing, variants in the CNV call but close to the ends of the CNV call will
be ignored. `margin.pct` defines the percentage of CNV length, located at 
each CNV limit, where variants will be ignored. For example, for a CNV 
chr1:1000-2000 and a `margin.pct` value of 10,
variants within chr1:1000-1100 and chr1:1900-2000 will be ignored.



# Plotting results

We can plot easily a certain CNV and the variants in it. For example,  the 
duplication CNV with `cnv.id`=17 can be plotted as follows:
```{r, fig.wide=TRUE}
plotVariantsForCNV(results, "16")
```

Some parameters can be customized, like `points.cex` and `points.pch`. It is 
also possible to plot all CNVs in a global schema where all the chromosomes are
plotted:
```{r, fig.width=8, fig.height=8, fig.wide=TRUE}
cnvs.file <- system.file("extdata", "DECoN.CNVcalls.2.csv",
                         package = "CNVfilteR", mustWork = TRUE)
cnvs.gr.2 <- loadCNVcalls(cnvs.file = cnvs.file, chr.column = "Chromosome",
                          start.column = "Start", end.column = "End", 
                          cnv.column = "CNV.type", sample.column = "Sample",
                          genome = "hg19")
plotAllCNVs(cnvs.gr.2)
```

Note that if a CNV is too small, it will not be visible when calling 
`plotAllCNVs()`.


# Session Info
```{r}
  sessionInfo()
```
  
